---
title: "Predicting the age of abalone"
author: "Abirham"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: true
---


## Methods
```{r, eval=FALSE, message=FALSE, warning=FALSE}
#Run these commands to install all the required packages for this report
install.packages("readr")
install.packages("dplyr")
install.packages("car")
install.packages("lmtest")
install.packages("ggplot2")
install.packages("GGally")
install.packages("gridExtra")
install.packages("MASS")
install.packages("leaps")
install.packages("glmnet")
install.packages("caret")
install.packages("gbm")

```

```{r Installing/Loading Packages, message=FALSE, warning=FALSE}
#loading all necessary packages 
library(readr)
library(dplyr)
library(car)
library(lmtest)
library(ggplot2)
library(GGally)
library(gridExtra)
library(MASS)
library(leaps)
library(glmnet)
library(caret)
library(gbm)
```



```{r}
# Corrected file path with forward slashes
abalone=read.csv("D:/Githubfordatamining/FA24-DATS6101-11/HW/Abalone_data.csv")
head(abalone)


```


```{r}
abalone$Sex <- as.factor(abalone$Sex)
str(abalone, give.attr = FALSE)
```

```{r}
library(knitr)
kable(abalone[1:10,], digits = 4,format = 'markdown')
```

```{r}
library(skimr)
skim(abalone)


```

```{r}

categoric_features <- colnames(abalone)[sapply(abalone, function(col) is.factor(col) || is.character(col))]

categoric_features

```

```{r}
numeric_features <- abalone %>% 
  select_if(is.numeric) %>% 
  colnames()
numeric_features
```
The dataset contains information on 4177 abalones, each described by 9 variables. Below is an overview of the dataset's structure:

Sex (Categorical): Indicates the gender of the abalone, represented as "M" (male), "F" (female), or "I" (infant).
Length (Numeric): The longest measurement of the abalone in millimeters.
Diameter (Numeric): The perpendicular width of the abalone.
Height (Numeric): The height of the abalone with the shell on.
Whole weight (Numeric): The weight of the whole abalone in grams.
Shucked weight (Numeric): The weight of the meat extracted from the abalone in grams.
Viscera weight (Numeric): The weight of the gut content in grams.
Shell weight (Numeric): The weight of the dried shell in grams.
Rings (Numeric): The number of growth rings observed in the shell, which serves as the dependent variable.


Looking at the data set above  summary, we can see that data is quite evenly distributed between the three factor levels of male, female and infant.

Also  we see that there are four different measure of weight i.e. Whole_weight, Shucked_weight, Viscera_weight and Shell.weight. Whole_weight is linear function of other weight predictors with Unknown mass of water/blood lost from shucking process.

Response variable:
The dependent variable, Rings, is included in the dataset and represents the number of rings observed after cutting and examining an abalone. While it does not directly indicate the age of an abalone, it serves as a reliable proxy. The age of an abalone can be calculated as Age=Rings + 1.5. Given the consistency of this relationship, Rings will be treated as the dependent variable in the analysis. The number of rings recorded in the dataset ranges from 1 to 29, with the majority of abalones having between 5 and 15 rings. The distribution is slightly positively skewed, but this does not present any significant challenges for the subsequent analysis.




```{r}
library(ggplot2)
library(gridExtra) 
library(grid)     

numeric_features <- colnames(abalone)[sapply(abalone, is.numeric)]
# Create a list to store individual ggplot objects
plots <- list()

# Generate histograms for each numeric feature
for (i in seq_along(numeric_features)) {
  num_feat <- numeric_features[i]
  p <- ggplot(abalone, aes_string(x = num_feat)) +
    geom_histogram(aes(y = ..density..), color = "gray", 
                   fill = scales::hue_pal()(length(numeric_features))[i], bins = 30) +
    geom_density(color = "black", size = 1) +
    labs(title = paste("Distribution of", num_feat),
         x = num_feat,
         y = "Density") +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  # Print each plot to check
  print(p)
  
  # Store the plot in the list
  plots[[i]] <- p
}

```
The distributions indicate that the majority of abalones have measurements and weights within moderate ranges. Most variables, particularly physical dimensions, exhibit symmetry, while weights and the response variable (rings) are slightly right-skewed. The Rings variable shows a central peak around 9–10 rings, with most values ranging between 5 and 15, and an overall range from 1 to 29. we can see that most abalones in the dataset are of moderate age. Further analysis can focus on understanding how these features influence the number of rings (age).

```{r}

group_1 <- c("Length", "Diameter", "Height","Shell.weight")
group_2 <- c("Whole.weight", "Shucked.weight","Viscera.weight" ,"Rings")

# First pairs plot
plot1 <- ggpairs(abalone[, c(group_1, "Sex")], 
                 aes(colour = Sex, alpha = 0.8),
                 title = "Pairs Plot - Group 1") +
  theme_grey(base_size = 8)

# Second pairs plot
plot2 <- ggpairs(abalone[, c(group_2, "Sex")], 
                 aes(colour = Sex, alpha = 0.8),
                 title = "Pairs Plot - Group 2") +
  theme_grey(base_size = 8)

# Display both plots
print(plot1)
print(plot2)


```

From the above pair plot we can see that their is  strong positive correlations between physical dimensions (Length, Diameter, Height) and Shell Weight, with larger abalones having heavier shells, and distributions varying slightly by sex.There is also strong correlations among Whole Weight, Shucked Weight, and Viscera Weight, while the correlation between Rings and these weights is moderate, particularly higher for "Infants" compared to "Males" and "Females." The Boxplots show that "Infants" have a wider spread in Rings, indicating more variability in age, whereas "Males" and "Females" show more consistent ranges.

```{r}

library(ggplot2)
for (num_feat in numeric_features) {
  p <- ggplot(data = abalone, aes_string(x = "Sex", y = num_feat)) +
    geom_boxplot(fill = "skyblue", outlier.color = "red", outlier.size = 1.5) + # Custom fill color
    labs(
      title = paste("Box Plot of", num_feat),
      x = "Sex",
      y = num_feat
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14)
    )
  
  print(p)
}


```
These box plots shows that Males and Females are generally larger and heavier than Infants across all dimensions and weights. The distribution of Rings shows that age is relatively consistent across sexes,while Infants show a slightly lower median and a narrower interquartile range, indicating that younger abalones tend to be smaller and there is some variation and more outliers among the adult groups. These observations suggest that sex is an important categorical factor to consider when analyzing relationships between features and age.


```{r}

library(ggplot2)
library(patchwork)
library(gridExtra)
library(ggplot2)

numeric_features <- colnames(abalone)[sapply(abalone, is.numeric)]

# Loop through each numeric feature and create Q-Q plots separately
for (num_feat in numeric_features) {
  # Generate the Q-Q plot
  p <- ggplot(data = abalone, aes(sample = .data[[num_feat]])) +
    stat_qq(color = "skyblue", size = 2) +  # Q-Q plot points
    stat_qq_line(color = "red", linetype = "dashed", size = 1) +  # Reference line
    labs(
      title = paste("Q-Q Plot of", num_feat),
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12)
    )
  
  print(p)
}

```
Most numeric features, except for Height and the weight variables, exhibit relatively normal behavior in the central range but deviate at the tails.

The Q-Q plot for Rings shows a step-like pattern, reflecting the discrete nature of the variable. While the central portion aligns with the normal distribution line, the tails deviate, particularly on the higher end.

These results suggest the need for potential transformations when modeling variables like weights and Rings to improve normality assumptions.

```{r}
library(ggplot2)
library(dplyr)

categoric_feature <- colnames(abalone)[sapply(abalone, is.factor)][1]


category_counts <- abalone %>%
  count(.data[[categoric_feature]]) %>%
  mutate(Percentage = n / sum(n) * 100)  

ggplot(category_counts, aes(x = "", y = Percentage, fill = .data[[categoric_feature]])) +
  geom_bar(stat = "identity", width = 1, color = "black") + 
  coord_polar(theta = "y", start = 0) +  
  scale_fill_brewer(palette = "Set3") +  
  labs(
    title = categoric_feature,
    fill = categoric_feature
  ) +
  theme_void(base_size = 14) + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", Percentage)),
    position = position_stack(vjust = 0.5),
    size = 4,
    color = "black"
  )


```

As it is indicated in the pie chart the distribution is fairly balanced across the three categories, with males slightly outnumbering the other two groups. from this we can say that sex-based analysis can be conducted without any major skew between groups.


```{r}
library(reshape2)

numeric_features <- colnames(abalone)[sapply(abalone, is.numeric)]
corr_matrix <- cor(abalone[, numeric_features], use = "complete.obs", method = "pearson")
mask <- upper.tri(corr_matrix)


corr_melt <- melt(corr_matrix)
corr_melt <- corr_melt[!mask, ]

ggplot(corr_melt, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "black", linewidth = 1) +  
  scale_fill_gradient2(low = "red", high = "skyblue", mid = "white", midpoint = 0) +  
  geom_text(aes(label = sprintf("%.2f", value)), size = 3.5, fontface = "bold", color = "black") + 
  labs(title = "Correlation Matrix", fill = "Correlation") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 16, margin = margin(b = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```

The correlation matrix shows that there is a strong interdependence among physical and weight features, while Rings is moderately correlated with size variables, particularly Shell Weight. These insights suggest that multicollinearity might need to be addressed when building predictive models, and Shell Weight could be a strong predictor of age