---
title: "Predicting the age of abalone"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Methods
```{r, eval=FALSE, message=FALSE, warning=FALSE}
#Run these commands to install all the required packages for this report
install.packages("readr")
install.packages("dplyr")
install.packages("car")
install.packages("lmtest")
install.packages("ggplot2")
install.packages("GGally")
install.packages("gridExtra")
install.packages("MASS")
install.packages("leaps")
install.packages("glmnet")
install.packages("caret")
install.packages("gbm")
install.packages("gplots")

```

```{r Installing/Loading Packages, message=FALSE, warning=FALSE}
#loading all necessary packages 
library(readr)
library(dplyr)
library(car)
library(lmtest)
library(ggplot2)
library(GGally)
library(gridExtra)
library(MASS)
library(leaps)
library(glmnet)
library(caret)
library(gbm)
library(gplots)
```

### Data summary and statistics

```{r, message=FALSE, warning=FALSE}
abalone <- read_csv("Abalone_data.csv")

is.factor(abalone$Sex)

abalone$Sex <- as.factor(abalone$Sex)
str(abalone, give.attr = FALSE)

library(knitr)
kable(abalone[1:10,], digits = 4,format = 'markdown')
```

##### Categorial Variables

| Sex | Number of observations |
| --- | ---------------------- |
|  M  | 1527 |
|  F  | 1307 |
|  I  | 1342 |

##### Numeric Variables

|      | Length | Diameter | Height | Whole Weight | Shucked Weight  | Viscera Weight | Shell Weight | Rings |
| ---- | ------ | -------- | ------ | ----- | ------- | ------- | ----- | ----- |
| Min  | 15.0   | 11.00    | 0.0000 | 0.4   | 0.20    | 0.10    | 0.30  | 1     |
|Median| 109.0  | 85.00    | 28.00  | 159.9 | 67.20   | 34.20   | 46.80 | 9     |
| Mean | 104.8  | 81.58    | 27.91  | 165.8 | 71.88   | 36.12   | 47.77 | 9.932 |
| Max  | 163.0  | 130.00   | 226.00 | 565.1 | 297.60  | 152.00  | 201.00| 29    |
| Cor  | 0.557  | 0.5750   | 0.5581 | 0.5408| 0.4212  | 0.5043  | 0.6280| 1.000 | 
 
 - Looking at the dataset summary, we can see that data is quite evenly distributed between the three factor levels of `male`, `female` and `infant`. 
 
 - Also from we see that there are four different measure of weight i.e. `Whole_weight`, `Shucked_weight`, `Viscera_weight` and `Shell.weight`. `Whole_weight` is linear function of other weight predictors with Unknown mass of water/blood lost from shucking process. Also we observed that min value of predictor `Height` is 0. Practically this is not possible, we will investigate these observations to look closely.
 
##### Response variable

- The dependent variable Rings is included in the dataset. It was measured as the number of rings observed after cutting and examining an abalone. Although it does not denote the age of a given abalone directly, it determines it more-or-less perfectly. The age of an abalone equals Rings + 1.5 . Since this relationship holds reliably, Rings will be considered as the dependent variable. The number of rings measured in the data ranges from 1 to 29 and most of the abalone have between 5 and 15 rings. The distribution is slightly positively skewed as well but this does not pose any specific problems for the further analysis. (see plot below)
####################################################################