---
title: "Predicting the age of abalone"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Methods
```{r, eval=FALSE, message=FALSE, warning=FALSE}
#Run these commands to install all the required packages for this report
install.packages("readr")
install.packages("dplyr")
install.packages("car")
install.packages("lmtest")
install.packages("ggplot2")
install.packages("GGally")
install.packages("gridExtra")
install.packages("MASS")
install.packages("leaps")
install.packages("glmnet")
install.packages("caret")
install.packages("gbm")
install.packages("gplots")

```

```{r Installing/Loading Packages, message=FALSE, warning=FALSE}
#loading all necessary packages 
library(readr)
library(dplyr)
library(car)
library(lmtest)
library(ggplot2)
library(GGally)
library(gridExtra)
library(MASS)
library(leaps)
library(glmnet)
library(caret)
library(gbm)
library(gplots)
```

### Data summary and statistics

```{r, message=FALSE, warning=FALSE}
abalone <- read_csv("Abalone_data.csv")

is.factor(abalone$Sex)

abalone$Sex <- as.factor(abalone$Sex)
str(abalone, give.attr = FALSE)

library(knitr)
kable(abalone[1:10,], digits = 4,format = 'markdown')
```

##### Categorial Variables

| Sex | Number of observations |
| --- | ---------------------- |
|  M  | 1527 |
|  F  | 1307 |
|  I  | 1342 |

##### Numeric Variables

|      | Length | Diameter | Height | Whole Weight | Shucked Weight  | Viscera Weight | Shell Weight | Rings |
| ---- | ------ | -------- | ------ | ----- | ------- | ------- | ----- | ----- |
| Min  | 15.0   | 11.00    | 0.0000 | 0.4   | 0.20    | 0.10    | 0.30  | 1     |
|Median| 109.0  | 85.00    | 28.00  | 159.9 | 67.20   | 34.20   | 46.80 | 9     |
| Mean | 104.8  | 81.58    | 27.91  | 165.8 | 71.88   | 36.12   | 47.77 | 9.932 |
| Max  | 163.0  | 130.00   | 226.00 | 565.1 | 297.60  | 152.00  | 201.00| 29    |
| Cor  | 0.557  | 0.5750   | 0.5581 | 0.5408| 0.4212  | 0.5043  | 0.6280| 1.000 | 
 
 - Looking at the dataset summary, we can see that data is quite evenly distributed between the three factor levels of `male`, `female` and `infant`. 
 
 - Also from we see that there are four different measure of weight i.e. `Whole_weight`, `Shucked_weight`, `Viscera_weight` and `Shell.weight`. `Whole_weight` is linear function of other weight predictors with Unknown mass of water/blood lost from shucking process. Also we observed that min value of predictor `Height` is 0. Practically this is not possible, we will investigate these observations to look closely.
 
##### Response variable

- The dependent variable Rings is included in the dataset. It was measured as the number of rings observed after cutting and examining an abalone. Although it does not denote the age of a given abalone directly, it determines it more-or-less perfectly. The age of an abalone equals Rings + 1.5 . Since this relationship holds reliably, Rings will be considered as the dependent variable. The number of rings measured in the data ranges from 1 to 29 and most of the abalone have between 5 and 15 rings. The distribution is slightly positively skewed as well but this does not pose any specific problems for the further analysis. (see plot below)
####################################################################

#EDA in here between pair plot and preprocessing
"
```{r}
abalone <- abalone %>% filter(Height > 0)

```

```{r}
summary(abalone)
```

```{r}
# Distribution of rings
ggplot(abalone, aes(x = Rings)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Rings", x = "Number of Rings", y = "Frequency")

```
####Observation: 
- The histogram shows most abalone have around 10 rings, indicating age. The distribution is right-skewed with fewer abalone having a high number of rings. Ring counts range from 0 to 30, decreasing in frequency with distance from 10.

```{r}
# Distribution of abalone sex (bar plot)
ggplot(abalone, aes(x = Sex, fill = Sex)) +
  geom_bar() +
  labs(title = "Distribution of Abalone Sex", x = "Sex", y = "Count") +
  scale_fill_manual(values = c("M" = "blue", "F" = "pink", "I" = "gray"))

```
#### Observation:
- The chart "Distribution of Abalone Sex" shows abalones categorized by sex: Female (~1300), Infant (~1200), and Male (~1500). This indicates a higher count of males followed by infants and female.

```{r}
# Whole weight vs Rings (scatterplot)
ggplot(abalone, aes(x = `Whole weight`, y = Rings, color = Sex)) +
  geom_point(alpha = 0.6) +
  labs(title = "Whole Weight vs Rings", x = "Whole Weight", y = "Rings")
```
#### Observation:
- The scatter plot shows that abalone ring count generally increases with whole weight, though there is considerable variability. Females, infants, and males are color-coded, showing their distribution. This indicates weight might be a significant predictor of age.


```{r}
# Height Distribution by sex (Box plot)
ggplot(abalone, aes(x = Sex, y = Height, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Height Distribution by Sex", x = "Sex", y = "Height")
```
#### Observation:
- The box plot shows the height distribution by sex (F: female, I: infant, M: male). It highlights the median, quartiles, and potential outliers for each category. Males tend to have slightly higher median heights compared to females and infants. This variation in height could be considered when predicting abalone age.


```{r}
# Whole weight vs Cobined Shucked and Viscera Weight (Scatter Plot)
ggplot(abalone, aes(x = `Whole weight`, y = `Shucked weight` + `Viscera weight`, color = Sex)) +
  geom_point(alpha = 0.6) +
  labs(title = "Whole Weight vs Combined Shucked and Viscera Weight", x = "Whole Weight", y = "Shucked + Viscera Weight")

```                                              
#### Observation:
- This scatter plot shows a positive correlation between whole weight and combined shucked plus viscera weight of abalones. Different colors represent the sexes: females (red), infants (green), and males (blue).This relationship suggests whole weight could help predict shucked weight.


```{r}
# Distribution of Rings by sex (Violin Plot)
ggplot(abalone, aes(x = Sex, y = Rings, fill = Sex)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  labs(
    title = "Distribution of Rings by Sex",
    x = "Sex",
    y = "Rings"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")

```

### Observation:
- The plot highlights the density of rings within each category. Males and females have similar distributions, while infants show a narrower spread. The variation in age across sexes is evident.

```{r}
abalone_no_sex <- abalone[, !names(abalone) %in% c("Sex","Infant")]

# Display the first few rows of the new dataset
head(abalone_no_sex)
```

```{r}
cor_matrix <- cor(abalone_no_sex)
print(cor_matrix)
```

```{r}

# Load the gplots library
library(gplots)

# Calculate the correlation matrix (exclude 'Sex' and 'Infant')
cor_matrix <- cor(abalone_no_sex)

# Draw the heatmap with correlation numbers
heatmap.2(cor_matrix, 
          main = "Heatmap of Correlation Matrix", 
          col = colorRampPalette(c("blue", "white", "red"))(50), # Color gradient
          scale = "none", # Don't scale the data
          trace = "none", # Remove the trace lines
          dendrogram = "none", # No clustering
          margins = c(8, 8), # Adjust margins for better label visibility
          cexRow = 0.8, # Font size for row labels
          cexCol = 0.8, # Font size for column labels
          cellnote = round(cor_matrix, 2), # Round correlation values to 2 decimal places
          notecol = "black", # Color of the numbers
          notecex = 0.8) # Size of the numbers

```